---
title: "QGP MEV Analysis"
author: "Mohamed Nadhir Djekidel"
output: html_document
---



# Load libraries

```{r}
suppressPackageStartupMessages(require(tidyverse))
suppressPackageStartupMessages(require(glue))
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(readr))
suppressPackageStartupMessages(require(VariantAnnotation))
suppressPackageStartupMessages(require(purrr))
suppressPackageStartupMessages(require(GenomicRanges))
suppressPackageStartupMessages(require(tidyverse))
suppressPackageStartupMessages(require(SCP))
suppressPackageStartupMessages(require(RColorBrewer))

source("../R/utils.R")
```


```{r}
dout="../data/MEGAnE_analysis"
fs::dir_create(dout)
```


## Load data

```{r}
QGP_population_structure = readr::read_tsv(
  "../data-raw/QGP_populations/qgp_population_structure_clusters_ids.txt",col_names = T)

head(QGP_population_structure)
```


```{r}
QGP_population_structure = readr::read_tsv("../data-raw/qgp_population_structure.txt", col_names = F,show_col_types = F)
colnames(QGP_population_structure) = c("Sample","Population")

QGP_population_structure$Population[ QGP_population_structure$Population == "F1" ] = "QGP_GA"
QGP_population_structure$Population[ QGP_population_structure$Population == "F2" ] = "QGP_AFR"
QGP_population_structure$Population[ QGP_population_structure$Population == "F3" ] = "QGP_WEP"
QGP_population_structure$Population[ QGP_population_structure$Population == "F4" ] = "QGP_PAR"
QGP_population_structure$Population[ QGP_population_structure$Population == "F6" ] = "QGP_SAS"
QGP_population_structure$Population[ QGP_population_structure$Population == "F9" ] = "QGP_ADM"

## we have only one sample classified as F7
QGP_population_structure = subset(QGP_population_structure, Population != "F7")


QGP_population_structure$Population = factor(QGP_population_structure$Population, 
                                             levels =  rev(names(sort(table(QGP_population_structure$Population))))
                                             )
```


```{r}
Sys.setenv(VROOM_CONNECTION_SIZE = 5000000)
QGP_ME_VCF <- VariantAnnotation::readVcf("../MeGAne/QGP_MeGANE_jointcall_out/QGP_common/QGP_biallelic_PASS_vcf.gz",genome="hg38")
dim(geno(QGP_ME_VCF)$GT)
```

```{r}
gp1000_ME_VCF <- VariantAnnotation::readVcf("../MeGAne/1000GP/phasing_out/GRCh38_2504/GRCh38_2504_biallelic_PASS.fixed.vcf.gz",
                                            genome="hg38")
dim(geno(gp1000_ME_VCF)$GT)
```

```{r}
gp1000_SVTYPES <- readr::read_tsv("../MeGAne/BBJ_Megane/hum0014.v28.MEs.v1/BBJ_4880_subjects_MEV_POS_AF.GRCh37.txt.gz",col_names = FALSE)
```

```{r}
BBJ_SVTYPES <- readr::read_tsv("../MeGAne/BBJ_Megane/hum0014.v28.MEs.v1/BBJ_4880_subjects_MEV_POS_AF.GRCh37.txt.gz",
                               col_names = TRUE,show_col_types = FALSE) |>
  dplyr::rename(SVTYPE = ME_CLASS, MEI = ME, CHROM=CHR)
```


## STATs about the distribution of the detected Mobile-elements

```{r}
QGP_SVTYPES = as.data.frame(VariantAnnotation::info(QGP_ME_VCF))|> 
  mutate(
    CHROM = as.character(seqnames(rowRanges(QGP_ME_VCF))),
    POS   = start(rowRanges(QGP_ME_VCF)),
    ID    = rownames(VariantAnnotation::info(QGP_ME_VCF))
  ) |>
  select(ID, CHROM, POS, SVTYPE, everything()) |> 
   dplyr::mutate(SVTYPE = unlist(SVTYPE), 
                 MEI = unlist(MEI),
                 MEPRED = sapply(MEPRED,function(x) {
                   ifelse(length(x) == 0, "",x)
                 }), 
                 MESTRAND = sapply(MESTRAND,function(x) {
                   ifelse(length(x) == 0, "+",x)
                 })
                 ) 

gp1000_SVTYPES = as.data.frame(VariantAnnotation::info(gp1000_ME_VCF)) |>
  mutate(
    CHROM = as.character(seqnames(rowRanges(gp1000_ME_VCF))),
    POS   = start(rowRanges(gp1000_ME_VCF)),
    ID    = rownames(VariantAnnotation::info(gp1000_ME_VCF))
  ) |>
  dplyr::select(ID, CHROM, POS, SVTYPE, everything()) |> 
  dplyr::mutate(SVTYPE = unlist(SVTYPE), 
                 MEI = unlist(MEI),
                 MEPRED = sapply(MEPRED,function(x) {
                   ifelse(length(x) == 0, "",x)
                 })
                 )

```


```{r}
gp1000_SVTYPES_unique = gp1000_SVTYPES |>
  dplyr::select(SVTYPE, MEI) |>
  dplyr::distinct()

QGP_SVTYPES_unique = QGP_SVTYPES  |>
  dplyr::select(SVTYPE, MEI) |>
  dplyr::distinct()

table(QGP_SVTYPES_unique$MEI %in% gp1000_SVTYPES_unique$MEI)
```

## Classify MEI into categories

```{r}
classify_part <- function(p) {
  if (grepl("^AluY", p))                                      return("AluY")
  if (grepl("^Alu", p) || p %in% c("FLAM_C","FLAM_A","FRAM",
                                     "FAM","FLA","FLAM","ALU")) return("Alu")
  if (p == "L1HS")                                             return("L1HS")
  if (grepl("^L1", p) || p %in% c("pA","pT"))                 return("L1")
  if (grepl("^SVA_", p))                                       return("SVA")
  if (p %in% c("LTR5_Hs","LTR5B","HERVK","HERVK3I","HERVK11I",
               "HERVKC4","HERV-K14I","MER11B","MER11C",
               "LTR13","LTR14"))                               return("HERV-K")
  return("Others")
}

tie_priority <- c("AluY","Alu","L1HS","HERV-K","L1","SVA","Others")
 
classify_mei <- function(mei) {
  parts  <- trimws(strsplit(as.character(mei), "\\|")[[1]])
  cats   <- sapply(parts, classify_part, USE.NAMES = FALSE)
  counts <- sort(table(cats), decreasing = TRUE)
 
  # Check for tie at the top
  top_count <- counts[1]
  tied_cats <- names(counts[counts == top_count])
 
  if (length(tied_cats) == 1) {
    return(tied_cats)
  } else {
    # Break tie by priority
    for (p in tie_priority) {
      if (p %in% tied_cats) return(p)
    }
  }
  return("Others")
}
 
# Apply classification
QGP_SVTYPES$category  <- sapply(QGP_SVTYPES$MEI,  classify_mei)
gp1000_SVTYPES$category <- sapply(gp1000_SVTYPES$MEI, classify_mei)

BBJ_SVTYPES$category <- sapply(BBJ_SVTYPES$MEI, classify_mei)

```

```{r}
100 * sort(table(QGP_SVTYPES$category),decreasing = T)/nrow(QGP_SVTYPES)
```


```{r}
100 * sort(table(gp1000_SVTYPES$category),decreasing = T)/nrow(gp1000_SVTYPES)
```

```{r}
100 * sort(table(BBJ_SVTYPES$category),decreasing = T)/nrow(BBJ_SVTYPES)
```


## Update SVTYPEs in QGP VCF

```{r}

pos = match(QGP_SVTYPES$SVTYPE , QGP_SVTYPES_unique$SVTYPE)
QGP_SVTYPES$SVTYPE_1000gp = QGP_SVTYPES_unique$SVTYPE_1000gp[pos]

if( all(rownames(QGP_SVTYPES) == info(QGP_ME_VCF) |> rownames())){
  info(QGP_ME_VCF)$SVTYPE =  QGP_SVTYPES$SVTYPE_1000gp 
}


info(QGP_ME_VCF)$METYPE= info(QGP_ME_VCF)$SVTYPE

info(QGP_ME_VCF)$SVTYPE= "INS"

hdr <- header(QGP_ME_VCF)
hdr_info = info(hdr)

new_field <- DataFrame(
  Number = ".",
  Type = "String",
  Description = "MEGAnE assigned SVTYPE",
  row.names = "METYPE"
)

hdr_info = rbind(hdr_info, new_field)

info(hdr) = hdr_info

header(QGP_ME_VCF) = hdr

QGP_ME_VCF = GenomicRanges::sort(QGP_ME_VCF)



VariantAnnotation::writeVcf(obj = QGP_ME_VCF, 
                            filename = "../MeGAne/QGP_MeGANE_jointcall_out/QGP_common/QGP_biallelic_PASS_Updated.vcf")
```


## Plot major MEV categories distribution across QGP, 1000 GP, and BBJ populations

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)



me_colors <- c(
  "AluY"   = "#ffcb69",
  "Alu"    = "#e8ac65",
  "L1HS"   = "#73bfb8",
  "SVA"    = "#3da5d9",
  "L1"     = "#2364aa",
  "HERV-K" = "#2c6e49",
  "Others" = "#BFBFBF" 
)


qgp_pie   <- QGP_SVTYPES |>
  dplyr::mutate(category = factor(category, levels = names(me_colors) )) |>
  dplyr::group_by(category) |>
  dplyr::tally() |>
  dplyr::mutate(pct=100 * n/sum(n)) |>
  dplyr::arrange(category) |>
  dplyr::mutate(label = glue("{category} ({round(pct,2)}%)"))
  
  

gp1000_pie <- gp1000_SVTYPES |>
  dplyr::mutate(category = factor(category, levels = names(me_colors) )) |>
  dplyr::filter(CHROM %in% unique(QGP_SVTYPES$CHROM)) |>
  dplyr::group_by(category) |>
  dplyr::tally() |>
  dplyr::mutate(pct=100 * n/sum(n)) |>
  dplyr::arrange(category) |>
  dplyr::mutate(label = glue("{category} ({round(pct,2)}%)"))


BBJ_pie <- BBJ_SVTYPES |>
  dplyr::mutate(category = factor(category, levels = names(me_colors) )) |>
  dplyr::group_by(category) |>
  dplyr::tally() |>
  dplyr::mutate(pct=100 * n/sum(n)) |>
  dplyr::arrange(category) |>
  dplyr::mutate(label = glue("{category} ({round(pct,2)}%)"))

```


```{r fig.height=6, fig.width=12}
p1 = plotthis::PieChart(qgp_pie,x="label", y="pct",
                        group_by="category",
                        label = "category",
                   #label = "category",
                        palcolor = me_colors, 
                   title = "QGP") +
  theme(legend.position = "bottom", 
        plot.title = element_text(hjust = 0.5))

p2 = plotthis::PieChart(gp1000_pie,x="label", y="pct",
                        group_by="category",
                        label = "category",
                        palcolor = me_colors, 
                        title = "1000 GP") +
  theme(legend.position = "bottom", 
        plot.title = element_text(hjust = 0.5))


p3 = plotthis::PieChart(BBJ_pie,x="label", y="pct",
                        group_by="category",
                        label = "category",
                        palcolor = me_colors, 
                        title = "BBJ") +
  theme(legend.position = "bottom", 
        plot.title = element_text(hjust = 0.5))


pdf(file =  "../data/MEGAnE_analysis/MEV_families_proportions_v2.pdf",width = 12,height = 6)
print(p2 + p1 + p3)
dev.off()

print(p2 + p1 + p3)
```


```{r}
subject_mapping = readr::read_tsv("../data-raw/006_subject_mapping.txt")
subject_clusters = readr::read_tsv("../data-raw/QGP_populations/qgp_population_structure_clusters_ids.txt")
```

```{r}
table(gsub("_.*","",QGP_population_structure$Sample) %in% subject_mapping$old_sidra_subject_id)
```

```{r}
QGP_population_structure$ID = gsub("_.*","",QGP_population_structure$Sample)
pos_new = which(QGP_population_structure$ID %in% subject_mapping$new_sidra_subject_id)
new_idx = match(QGP_population_structure$ID[pos_new] , subject_mapping$new_sidra_subject_id)

pos_old = which(QGP_population_structure$ID %in% subject_mapping$old_sidra_subject_id)
old_idx = match(QGP_population_structure$ID[pos_old] , subject_mapping$old_sidra_subject_id)

QGP_population_structure$NewID = NA

QGP_population_structure$NewID[pos_new] = subject_mapping$new_sidra_subject_id[new_idx]
QGP_population_structure$NewID[pos_old] = subject_mapping$new_sidra_subject_id[old_idx]
```


```{r}
table(  QGP_population_structure$NewID  %in%  gsub("_.*","",QGP_population_structure$Sample) ) 
```

```{r}
setdiff(  gsub("_.*","",geno(QGP_ME_VCF)$GT |> colnames())  ,  gsub("_.*","",QGP_population_structure$NewID) ) |> head()
```

```{r}
readr::write_tsv(QGP_population_structure, file="../data/MEGAnE_analysis/QGP_population_structure_final.tsv")
```

```{r}
head(QGP_population_structure)
```


## Load PCA analysis results

```{r}
KGP_PCA = readr::read_tsv("../MeGAne/PCA_analysis/06_pca/KGP_ref.eigenvec")
colnames(KGP_PCA) = gsub("#FID","FID",colnames(KGP_PCA))

QGP_PCA = readr::read_tsv("../MeGAne/PCA_analysis/06_pca/QGP.pruned.pruned.eigenvec")
colnames(QGP_PCA) = gsub("#FID","FID",colnames(QGP_PCA))

QGP_PCA$IID = gsub("^1_","",QGP_PCA$IID)
QGP_PCA$FID = gsub("^1_","",QGP_PCA$FID)
QGP_PCA$ID = gsub("_.*","",QGP_PCA$IID)

pos = match(QGP_PCA$ID, QGP_population_structure$NewID)
QGP_PCA$Population= QGP_population_structure$Population[pos]


## Assign colors to populations
pop_map <- tribble(
  ~Population,              ~Color,       ~code, ~qgp_pop,  ~hex,
  "AdmixedPop",      "Mushroom",    "00",  "QGP_ADM", "#4E1594",
  "GeneralArabs",    "Blueberry",   "Q1b", "QGP_GA", "#4E79A7",
  "EastAfricans",    "Orange",      "Q3",  "QGP_AFR", "#F28E2B",
  "PeninsularArabs", "Coral",       "Q1a", "QGP_PAR", "#76B7B2",
  "Persians",        "Strawberry",  "Q2a", "QGP_WEP", "#E15759",
  "South Asians",    "Lemon",       "Q2b", "QGP_SAS", "#EDC948"
)

pop_col_map = pop_map$hex
names(pop_col_map) =  pop_map$qgp_pop

QGP_PCA = QGP_PCA |> 
  dplyr::filter(!is.na(Population)) |>
  dplyr::left_join(pop_map)

plt_pca = plotthis::DimPlot(QGP_PCA,
                  dims = 3:4,
                  group_by = "Population",
                  label = TRUE,
                  label_insitu = TRUE,
                  label_bg = NA,
                  pt_size = 1,
                  pt_alpha = 0.6,
                  label_pt_color = "black",
                  label_bg_r = 0,
                  label_fg = "black",
                  label_size = 4,
                  label_repel = TRUE) +
  scale_color_manual(values = pop_col_map)

ggsave(plot = plt_pca, filename = "../data/MEGAnE_analysis/QGP_ME_PCA.pdf",width = 6,height = 6)
ggsave(plot = plt_pca, filename = "../data/MEGAnE_analysis/QGP_ME_PCA.png",width = 6,height = 6)
plt_pca
```

# Classify QGP allele frequencies into bins and annotate with SV types

```{r}
qgp_af = readr::read_tsv("../MeGAne/PCA_analysis/additional/QGP_AF.tsv",
                         col_names = FALSE)

colnames(qgp_af) = c("ID","AC","AN","AF")

qgp_af_classified <- qgp_af |>
  mutate(
    af_bin = case_when(
      AC == 1                  ~ "Singleton",
      AF > 0   & AF <  0.01   ~ "<1%",
      AF >= 0.01 & AF < 0.50  ~ "≥1%",
      AF >= 0.50               ~ "≥50%"
    ) |> factor(levels = c("≥50%", "≥1%", "<1%", "Singleton") |> rev())
  ) |>
  filter(!is.na(af_bin))
```


```{r}
qgp_af_classified_detailed = qgp_af_classified
pos = match(qgp_af_classified$ID, QGP_SVTYPES$ID)
qgp_af_classified_detailed$Class = QGP_SVTYPES$category[pos]
qgp_af_classified_detailed$ME_TYPE = ifelse(grepl("_ins_",qgp_af_classified_detailed$ID),"ME_insertion","ME_absence")
```

```{r}
qgp_gt_matrix <- geno(QGP_ME_VCF)$GT

qgp_gt_dosage = matrix(0, 
                   nrow = nrow(qgp_gt_matrix), 
                   ncol = ncol(qgp_gt_matrix),
                   dimnames = dimnames(qgp_gt_matrix)
                   )

qgp_gt_dosage[qgp_gt_matrix %in% c("0/0","0|0")] = 0
qgp_gt_dosage[qgp_gt_matrix %in% c("0/1", "1/0","1/.","./1","1|0","0|1")] = 1
qgp_gt_dosage[qgp_gt_matrix %in% c("1/1","1|1")] = 2


qgp_smp_order = colSums(qgp_gt_dosage) |> order()

qgp_gt_dosage = qgp_gt_dosage[,qgp_smp_order]



cli::cli_progress_bar(total  = ncol(qgp_gt_dosage))

qgp_cum_stats = data.frame()
processed_me = c()

for(smp in colnames(qgp_gt_dosage)){
  
  cli::cli_progress_update()
  me_in_sample = rownames(qgp_gt_dosage)[which(qgp_gt_dosage[,smp] >0)]
  me_in_sample = setdiff(me_in_sample,processed_me)
  
  if(length(me_in_sample)>0){
    df = qgp_af_classified |>
      dplyr::filter(ID %in% me_in_sample) |>
      dplyr::rename(ME_freq = af_bin) |>
      dplyr::group_by(ME_freq) |>
      tally() |>
      dplyr::mutate(Sample = smp) |>
      dplyr::select(Sample, ME_freq, n) |>
      tidyr::complete(
          Sample,
          ME_freq,              # uses factor levels
          fill = list(n = 0)
        ) |>
      pivot_wider(
        names_from  = ME_freq,
        values_from = n,
      ) |>
      column_to_rownames("Sample")
    
    if(nrow(qgp_cum_stats)>0){
        df2 = rbind(qgp_cum_stats[nrow(qgp_cum_stats),],df)
        cs = colSums(df2)
        df[1,] = cs
    }
        
    qgp_cum_stats = rbind(qgp_cum_stats, df)
    processed_me = c(processed_me,me_in_sample)
  }
}
cli::cli_progress_done()
```


```{r}
qgp_cum_stats$Step = 1:nrow(qgp_cum_stats)

qgp_cum_stats_long = qgp_cum_stats |>
  pivot_longer(!Step,names_to = "AF_bin",values_to = "cumSum") |>
  dplyr::mutate(AF_bin = factor(AF_bin,levels = c("≥50%", "≥1%", "<1%", "Singleton") |> rev())) |>
  dplyr::mutate(Sample = gsub("_.*","",colnames(qgp_gt_dosage)[Step])) |>
  dplyr::left_join(QGP_population_structure[,c("NewID","Population")],by = c("Sample" = "NewID"))
```


```{r fig.height=6, fig.width=6}

af_colors <- 
c(
  "Singleton" = "#f5e6e8",
  "<1%"       = "#d5c6e0",
  "≥1%"       = "#aaa1c8",
  "≥50%"      = "#192a51"
)

qgp_cum_stats_long = subset(qgp_cum_stats_long, !is.na(qgp_cum_stats_long$Population))

qgp_cum_stats_plt = ggplot(qgp_cum_stats_long,aes(x=Step,y=cumSum, fill=AF_bin)) + 
  geom_area() + 
  geom_rug(aes(color=Population,group = Population),sides="b", position = "jitter") +
  scale_fill_manual(values = af_colors) +  
  scale_color_manual(values = pop_col_map) +
  scale_x_continuous(
      breaks = scales::breaks_width(1000),
      labels = scales::label_comma()
    ) +
  theme_scp() +
  xlab(glue("QGP individuals (n={prettyNum(ncol(qgp_gt_matrix),big.mark = ',')})")) +
  ylab("Number of MEVs") +
  theme(aspect.ratio = 0.6,
        axis.title = element_text(size = 11),
        axis.text  = element_text(size = 9) )

ggsave(plot = qgp_cum_stats_plt, filename = "../data/MEGAnE_analysis/QGP_ME_cumFreq.pdf",width = 6,height = 6)
ggsave(plot = qgp_cum_stats_plt, filename = "../data/MEGAnE_analysis/QGP_ME_cumFreq.png",width = 6,height = 6)

qgp_cum_stats_plt
```

## Generate ME families distribution by QGP population

```{r fig.height=12, fig.width=12}
QGP_populations_svtypes = list()

QGP_populations_svtypes_plts = list()

for(pop in sort(levels(QGP_population_structure$Population))){
  
  
  pop_samples = QGP_population_structure |>
    dplyr::filter(Population == pop) |>
    pull(NewID)
  
  pos = match(pop_samples, gsub("_.*","", colnames(qgp_gt_dosage)))
  
  pos = pos[!is.na(pos)]
  
  rs = rowSums(qgp_gt_dosage[,pos]>0)
  
  pop_mevs = rownames(qgp_gt_dosage)[rs>0]
  
  pop_svtypes = QGP_SVTYPES[pop_mevs,]
  
  QGP_populations_svtypes[[pop]] = pop_svtypes
  
  
  pop_pie   <- pop_svtypes |>
          dplyr::mutate(category = factor(category, levels = names(me_colors) )) |>
          dplyr::group_by(category) |>
          dplyr::tally() |>
          dplyr::mutate(pct=100 * n/sum(n)) |>
          dplyr::arrange(category) |>
          dplyr::mutate(label = glue("{category} ({round(pct,2)}%)"))
  
  pop_plt = plotthis::PieChart(pop_pie,
                          x="label", 
                          y="pct",
                          group_by="category",
                          label = "category",
                          palcolor = me_colors, 
                          title = pop) +
    theme(legend.position = "right", 
          plot.title = element_text(hjust = 0.5,size = 11,face = "bold"), 
          legend.key.spacing.y = unit(0.15, "cm")) +
    labs(
      fill  = "ME family"
      )

  
  QGP_populations_svtypes_plts[[pop]] = pop_plt
}

QGP_populations_svtypes_plts_combined = patchwork::wrap_plots(QGP_populations_svtypes_plts)


ggsave(plot = QGP_populations_svtypes_plts_combined, 
       filename = "../data/MEGAnE_analysis/MEV_families_byPopulation.pdf",width = 12,height = 12)

ggsave(plot = QGP_populations_svtypes_plts_combined, 
       filename = "../data/MEGAnE_analysis/MEV_families_byPopulation.png",width = 12,height = 12)


writeListToExcelSheet(QGP_populations_svtypes,
                      outputFilePath = "../data/MEGAnE_analysis/MEV_families_byPopulation.xlsx")

QGP_populations_svtypes_plts_combined
```

## Cumulative distribution by ME family

```{r}
QGP_SVTYPES$category = factor(QGP_SVTYPES$category,levels = names(me_colors))

pop_cum_stats_all = list()

for(pop in sort(levels(QGP_population_structure$Population))){
  cli::cli_h2(pop)

  pop_samples = QGP_population_structure |>
    dplyr::filter(Population == pop) |>
    pull(NewID)
  
  pos = match(pop_samples, gsub("_.*","", colnames(qgp_gt_dosage)))
  
  pos = pos[!is.na(pos)]
  
  rs = rowSums(qgp_gt_dosage[,pos]>0)
  
  pop_mevs = rownames(qgp_gt_dosage)[rs>0]
  
  pop_gt_dosage = qgp_gt_dosage[pop_mevs,pos]
  
  pop_smp_order = colSums(pop_gt_dosage) |> order()
  pop_gt_dosage = pop_gt_dosage[,pop_smp_order]
  
  pop_cum_stats = data.frame()
  pop_processed_me = c()
  
  pop_svtypes = QGP_SVTYPES[pop_mevs,]
  
  cli::cli_progress_bar(total  = ncol(pop_gt_dosage))

  for(smp in colnames(pop_gt_dosage)){
  
   cli::cli_progress_update()
    me_in_sample = rownames(pop_gt_dosage)[which(pop_gt_dosage[,smp] >0)]
    me_in_sample = setdiff(me_in_sample,pop_processed_me)
    
    if(length(me_in_sample)>0){
      
       df = pop_svtypes |>
        dplyr::filter(ID %in% me_in_sample) |>
        dplyr::group_by(category) |>
        tally() |>
        dplyr::mutate(Sample = smp) |>
        dplyr::select(Sample, category, n) |>
        tidyr::complete(
            Sample,
            category,              # uses factor levels
            fill = list(n = 0)
          ) |>
        pivot_wider(
          names_from  = category,
          values_from = n,
        ) |>
        column_to_rownames("Sample")
       
       if(nrow(pop_cum_stats)>0){
            df2 = rbind(pop_cum_stats[nrow(pop_cum_stats),],df)
            cs = colSums(df2)
            df[1,] = cs
       }
            
       pop_cum_stats = rbind(pop_cum_stats, df)
       pop_processed_me = c(pop_processed_me,me_in_sample)
    }
  }
  cli::cli_progress_done()
  
  pop_cum_stats$Population = pop
  
  pop_cum_stats_all[[pop]] = pop_cum_stats
}
```

```{r fig.height=12, fig.width=12}
pop_cum_stats_plts = list()
pop_nb_mevs_all = data.frame()


for( pop in names(pop_cum_stats_all)){
  
  pop_cum_stats = pop_cum_stats_all[[pop]]
  
  pop_cum_stats$Step = 1:nrow(pop_cum_stats)
  
  pop_cum_stats_long = pop_cum_stats |>
    pivot_longer(!c(Step,Population),names_to = "MEV",values_to = "cumSum") |>
    dplyr::mutate(MEV = factor(MEV,levels = names(me_colors) |> rev())) 
  
  pop_nb_mev = sum(tail(pop_cum_stats[,names(me_colors)],1))
  
  df = tail(pop_cum_stats,1)
  df$total = pop_nb_mev
  
  pop_nb_mevs_all = rbind(pop_nb_mevs_all, df)

  pop_plt = ggplot(pop_cum_stats_long,aes(x=Step,y=cumSum, fill=MEV)) + 
    geom_area() + 
    scale_fill_manual(values = me_colors) +  
    scale_x_continuous(
        breaks = scales::breaks_width(1000),
        labels = scales::label_comma()
      ) +
    theme_scp() +
    xlab("") +
    ylab("Number of MEVs") +
    ggtitle(pop, subtitle = glue("{prettyNum(pop_nb_mev,big.mark = ',')} MEVs")) +
    theme(aspect.ratio = 0.6,
          plot.title = element_text(hjust = 0.5,size = 11),
          plot.subtitle = element_text(hjust = 0.5,size = 9),
          axis.title = element_text(size = 11),
          axis.text  = element_text(size = 9) )
  
  pop_cum_stats_plts[[pop]] = pop_plt
}


pop_cum_stats_plts_combined = patchwork::wrap_plots(pop_cum_stats_plts,guides = "collect")

ggsave(plot = pop_cum_stats_plts_combined, 
       filename = "../data/MEGAnE_analysis/MEV_families_cumulative_byPopulation.pdf",width = 12,height = 12)

ggsave(plot = pop_cum_stats_plts_combined, 
       filename = "../data/MEGAnE_analysis/MEV_families_cumulative_byPopulation.png",width = 12,height = 12)

pop_cum_stats_plts_combined
```


```{r fig.height=5, fig.width=5}
pop_nb_mevs_all_long = pop_nb_mevs_all |>
          dplyr::arrange(-total) |>
          dplyr::select(names(me_colors), Population) |>
          pivot_longer(!Population, names_to = "MEV",values_to = "Count") |>
          dplyr::mutate(MEV = factor(MEV, levels = names(me_colors) |> rev() )) |>
          dplyr::mutate(Population = factor(Population, levels = unique(Population))) 

pop_nb_mevs_plt = plotthis::BarPlot(pop_nb_mevs_all_long, x="Population",
                  y="Count",group_by = "MEV",
                  x_text_angle = 40
                  ) +
  scale_fill_manual(values = me_colors) +
  coord_cartesian(ylim = c(0,max(pop_nb_mevs_all$total * 1.1)))


ggsave(plot = pop_nb_mevs_plt, 
       filename = "../data/MEGAnE_analysis/nb_MEVs_byPopulation.pdf",width = 5,height = 5)

ggsave(plot = pop_nb_mevs_plt, 
       filename = "../data/MEGAnE_analysis/nb_MEVs_byPopulation.png",width = 5,height = 5)

writeExcelSheet(pop_nb_mevs_all_long, sheetName = "nb_MEVs_byPopulation",
                outputFilePath = "../data/MEGAnE_analysis/nb_MEVs_byPopulation.xlsx")

pop_nb_mevs_plt
```